Species richness destabilizes trophic communities in freshwater ecosystems

### 2020-03-25 manuscript/bef_stability/manuscript.Rmd
Improve SEM plot and the other Figures

---
title: Species richness destabilizes trophic communities in freshwater ecosystems
author: Alain Danet, Maud Mouchet, Elisa Thébault and Colin Fontaine
date: \today 
output:
  pdf_document:
    fig_caption: yes
fontsize: 12pt
header-includes:
   - \usepackage{setspace}
   - \doublespacing
   - \usepackage{lineno}
   - \linenumbers
geometry: margin=2cm
bibliography: references.bib
---

@@

We computed the connectance ($C = L/N^2$, L and N being resp. the number of
links and of nodes) and the average trophic level of each network, the average
trophic level was weighted the biomass of each node (excluding resource nodes
for which we had no biomass) ($\overline{T} = \sum_i^N{B_i * T_i} \times
\sum_i^N{b_i}^{-1}$, $T_i$ and $B_i$ being resp. the trophic level and the
biomass of the node $i$). In order to have connectance at the species level
(i.e. not a trophic species level), we merge the trophic species of the network
by species and thereby aggregating the link by species.

@@

p_format <- "png"
p_width <- 700
p_height <- 700

@@

graph_stab_sem <- build_diag_from_sem(stab_sem_rich, p_val_thd = 0.05)
render_graph(graph_stab_sem)
export_graph(graph = graph_stab_sem, file_type = p_format,
  file_name = stab_sem_path,
  height = p_height - p_height*20/100,
  width = p_width
)
#source(mypath("R", "plot_methods.R")) 
graph_bm_sem <- build_diag_from_sem(bm_sem_rich, p_val_thd = 0.05)
render_graph(graph_bm_sem)
export_graph(graph = graph_bm_sem, file_type = p_format,
  file_name = bm_sem_path,
  #file_name = "stab_sem.svg",
  height = p_height,
  width = p_width 
)

library(magick)
library(rsvg)
img_stab_sem <- stab_sem_path %>%
  image_read()
p_stab_sem_rich <- ggdraw() +
  draw_image(img_stab_sem)
img_bm_sem <- bm_sem_path %>%
  image_read()
p_bm_sem_rich <- ggdraw() +
  draw_image(img_bm_sem)

@@

```{r sem-plot, fig.dim = c(10,10), fig.cap = "SEM"}
#https://tex.stackexchange.com/questions/20496/drawing-different-tikz-shapes-parameterized-by-data-from-a-file?rq=1
#https://tex.stackexchange.com/questions/83888/how-to-plot-data-from-a-csv-file-using-tikz-and-csvsimple

plot_grid(p_stab_sem_rich, p_bm_sem_rich, labels = LETTERS[1:2], rel_widths = c(.55, .45))
```

@@

Species richness and the average trophic level were the main direct drivers of
total biomass ($r_\delta = `r rich_on_bm`$ and $r_\delta = `r tlvl_on_bm`$ resp.) whereas
connectance had no significant effect (Fig. \@ref(fig:sem-plot) C). The stream
size had a direct positive effect on total biomass ($r_\delta = `r RC1_on_bm`$) whereas
avg temperature, avg DBO, CV of river size and CV of DBO had no direct effects
on total biomass.

Species richness had also a indirect positive effect on total biomass through
average trophic level, species richness had a total positive effect of $`r rich_on_bm`$
on total biomass. The stream size had a total effect on total biomass as strong
as species richness ($r_\delta =`r env_on_bm["log_RC1"]`$), indirectly through average trophic level
and species richness. The avg temperature of the river had an indirect positive
effect on total biomass ($r_\delta = `r env_on_bm["log_RC2"]`$), through
average trophic level and species richness as well. Avg DBO, CV DBO and CV of
river size had no or weak indirect effect on total biomass
(resp. $r_\delta =`r env_on_bm["log_RC3"]`$, $r_\delta =`r env_on_bm["log_RC5"]`$, $r_\delta =`r env_on_bm["log_RC4"]`$).

@@

$\overline{CV_{sp}}$ ($r_\delta = `r cv_sp_on_stab`$) decreased biomass stability more than
synchrony ($r_\delta = `r sync_on_stab`$, Figure \@ref(fig:sem-plot)). Species richness had a strong
positive effect on $\overline{CV_{sp}}$ and a strong negative effect on synchrony
whereas average trophic level and connectance had no significant effects on either
$\overline{CV_{sp}}$ nor synchrony. The avg river size had a small positive direct
effect on synchrony ($r_\delta = `r RC1_on_sync`$). The other environmental variables had
no significant direct effects on either $\overline{CV_{sp}}$ nor synchrony.

@@

```{r}
richness_lab <- "Total richness (m2^-1) (Log 10)"
stab_lab <- "Biomass stability (Log 10)"
bm_lab <- "Total biomass (g/m2) (Log 10)"

@@

p_stab_div <- sem_data %>%
  ggplot(aes(x = log_rich_tot_std, y = log_stab_std))+
  geom_point() +
  ylim(lim_log_stab) +
  xlim(lim_log_rich_tot) +
  labs(x = richness_lab, y = stab_lab) +
  geom_line(data = stab_div_pred) +
  geom_ribbon(data = stab_div_pred, aes(ymin = conf.low, ymax = conf.high), fill
  = "grey20", alpha = .2) +
  annotate("text", x = 1.0, y = .8, label = make_label_rsq(rsq = stab_div_sum),
    parse = TRUE)

# Right plot
p_stab_div_troph <- stab_div_troph %>%
  ggplot(aes(x = log_rich_tot_std, y = log_stab_std, color = troph_group)) +
  geom_point() +
  ggplot2::scale_color_manual(
    labels = troph_group_labeller(),
    values = c("blue", "green", "red"),
    name = "Trophic group"
  ) +
  ylim(lim_log_stab) +
  xlim(lim_log_rich_tot) +
  geom_line(data = stab_troph_pred) +
  geom_ribbon(data = stab_troph_pred,
    aes(ymin = conf.low, ymax = conf.high, color = NULL, group = troph_group),
    alpha = .2) +
  annotate("text", x = 0.8, y = .8, label = make_label_rsq(rsq = stab_troph_sum),
    parse = TRUE) +
  #scale_x_log10() + scale_y_log10() +
  labs(x = richness_lab, y = stab_lab)

p_stab <- plot_grid(p_stab_div, p_stab_div_troph,
  ncol = 2, rel_widths = c(.4, .6),
  labels = c("A", "B") 
)
```

@@

p_bm_div <- sem_data %>%
  ggplot(aes(x = log_rich_tot_std, y = log_bm_std))+
  geom_point() +
  ylim(lim_log_bm) +
  xlim(lim_log_rich_tot) +
  labs(x = richness_lab, y = bm_lab) +
  geom_line(data = bm_div_pred) +
  geom_ribbon(data = bm_div_pred, aes(ymin = conf.low, ymax = conf.high), fill
  = "grey20", alpha = .2) +
  annotate("text", x = .8, y = 1.5, label = make_label_rsq(rsq = bm_div_sum),
    parse = TRUE)

p_bm_div_troph <- stab_div_troph %>%
  ggplot(aes(x = log_rich_tot_std, y = log_bm_std, color = troph_group)) +
  geom_point() +
  ylim(lim_log_bm) +
  xlim(lim_log_rich_tot) +
  ggplot2::scale_color_manual(
    labels = troph_group_labeller(),
    values = c("blue", "green", "red"),
    name = "Trophic group"
  ) +
  geom_line(data = bm_troph_pred) +
  geom_ribbon(data = bm_troph_pred,
    aes(ymin = conf.low, ymax = conf.high, color = NULL, group = troph_group),
    alpha = .2) +
  annotate("text", x = 0.8, y = 1.5, label = make_label_rsq(rsq = bm_troph_sum),
    parse = TRUE) +
  #scale_x_log10() + scale_y_log10() +
  labs(x = richness_lab, y = bm_lab)

p_bm <- plot_grid(p_bm_div, p_bm_div_troph, ncol = 2, rel_widths = c(.4, .6),
  labels = c("C", "D"))
```

```{r psbm, fig.dim = c(10,5)}
plot_grid(p_stab, p_bm, ncol = 1)

@@

---
# Total biomass versus species richness
---
species richness increased total biomass (`r b_co_r`% `r b_ci_r` by 1% species richness increase). Total
biomass was `r b_co_t`% `r b_ci_t` higher in the high than in the low trophic
group. However, the increase of total biomass with the increase of species
richness was lower in the high than the low trophic group (`r b_co_r_t3`% `r
b_ci_r_t3` and `r b_co_r_t2`% `r b_ci_r_t2` respectively,  by 1% species
richness increase).

@@

We found that species richness increased total biomass as found in salt
marshes, forests [@wu_relationship_2014,@shuaifeng_positive_2018]. But opposite
relation has been found in richness in grasslands [@grace_integrative_2016],
notably because of light competition. Precedent researches in plant communities
showed that the total biomass was primary driven by environmental drivers. In
contrast, we found that the species richness was one of the main driver of the
total biomass, together with the stream size and increasing avg temperature
(with decreasing altitude). The big effect of species richness is 
partly explained by its effect on food web assembly and so, avg trophic level
(`r round(.2/0.54 * 100)` % of richness effect on total biomass). The direct
positive effect of species richness on total biomass might be due to resource use
complementarity [@cardinale_species_2002] [add more direct REF]. It was however
unexpected that connectance had no effect on total biomass, because we could
imagine that connectance could also improve resource use, but this might be
true only under the assumption that all species have the same assimilation
efficiency.

We found that contrary to stream size, the effect of altitude of the river and
its avg temperature on total biomass was only indirect through community
structure (avg trophic level and connectance).

@@

The total effect of river size on biomass stability was as negative as species
richness, indicating that big rivers had a high negative effect on stability (TO MOVE).
We found a mix of both expected and unexpected results on the effects on
environment on community structure. First, we expected that bigger and low
altitude streams had higher species richness because they might receive more
confluents streams and so more species through immigration.
Stream size increased both connectance and avg trophic level with the same
magnitude. An increase in mean trophic position of piscivorous species in
bigger rivers could explain this relation [@post_ecosystem_2000]. In our
network reconstruction, the fish-fish interactions are more resolved than
fish-resource interactions, which might explain positive effect of stream size
on connectance. @petchey_predicting_2010 predicted that temperature should
increase connectance but we did not find any relationship but temperature (and
low altitude) increased avg trophic level (**Bullshit**). 
CV of enrichment decreased both connectance and avg trophic level, while avg
enrichment (and CV of flow) increased both. The latter might by explained by
the fact that enrichment can benefit to top trophic level
[@hulot_functional_2000]. One explanation for the former is that enrichment
variation might destabilize top predators (REF or Bullshit again).

@@

In turn, we found that synchrony and $\overline{CV_{sp}}$ both decreased biomass
stability, as predicted by the formulation of stability based on the CV
[@thibaut_understanding_2013]. However, it was unexpected that the average
population $\overline{CV_{sp}}$ had a stronger negative effect on biomass
stability than synchrony [@bluthgen_land_2016; @olivier_independent_nodate]. We
can advance two hypothesis that can raise this effect. First, the opportunity
for diet differentiation can be low in streams compare to other organisms. Bird
species
are for example quite specialized in seed, flying or soil arthropods or mammals
which gives opportunities for independent dynamics. Plants also can achieve
independent dynamics either spatial segregation or niche differentiation. Because
fish species even piscivorous one are quite opportunistic in their foraging.
Secondly, in trophic communities, species directly affect each other through
feeding interactions, and so could raise synchrony between species
[@raimondo_interspecific_2004] more than in systems where interactions happen
more in the form of competitive exploitation.

>>>>

---
title: Species richness destabilizes trophic communities in freshwater ecosystems
author: Alain Danet, Maud Mouchet, Elisa Thébault and Colin Fontaine
date: \today 
output:
  pdf_document:
    fig_caption: true 
    keep_tex: true
fontsize: 12pt
header-includes:
   - \usepackage{setspace}
   - \doublespacing
   - \usepackage{lineno}
   - \linenumbers
geometry: margin=2cm
bibliography: references.bib
---

@@

We computed the connectance ($C = L/N^2$, L and N being resp. the number of
links and of nodes) and the average trophic level of each network, the average
trophic level was weighted the biomass of each node (excluding resource nodes
for which we had no biomass) ($\overline{T} = \sum_i^N{B_iT_i \times
\sum_i^N{B_i}^{-1}}$, $T_i$ and $B_i$ being resp. the trophic level and the
biomass of the node $i$). In order to have connectance at the species level
(i.e. not a trophic species level), we merge the trophic species of the network
by species and thereby aggregating the link by species.

@@

p_format <- "png"
p_width <- 700*2
p_height <- 900*2

@@

graph_stab_sem <- build_diag_from_sem(
  stab_sem_rich, 
  p_val_thd = 0.05,
  env_x_pos_range = c(from = 0, to = 7.5),
  env_y_pos = 0,
  env_y_sep = 1.5,
  order_env_node = c("RC4", "log_RC3", "RC5", "log_RC1", "log_RC2")  
)
graph_stab_sem %<>%
  add_global_graph_attrs(
    attr = "splines",
    value = "ortho", #spline
    attr_type = "graph")
render_graph(graph_stab_sem)
export_graph(graph = graph_stab_sem, file_type = p_format,
  file_name = stab_sem_path,
  height = p_height - p_height*30/100,
  width = p_width
)
#source(mypath("R", "plot_methods.R")) 
graph_bm_sem <- build_diag_from_sem(
  bm_sem_rich, 
  p_val_thd = 0.05,
  env_x_pos_range = c(from = 0, to = 7.5),
  env_y_pos = 0,
  env_y_sep = 1.5,
  order_env_node = c("RC4", "log_RC3", "RC5", "log_RC1", "log_RC2")  
)
graph_bm_sem %<>%
  add_global_graph_attrs(
    attr = "splines",
    value = "ortho", #spline
    attr_type = "graph")
render_graph(graph_bm_sem)
export_graph(graph = graph_bm_sem, file_type = p_format,
  file_name = bm_sem_path,
  height = p_height - p_height*45/100,
  width = p_width 
)

library(magick)
library(rsvg)
img_stab_sem <- stab_sem_path %>%
  image_read()
p_stab_sem_rich <- ggdraw() +
  draw_image(img_stab_sem)
img_bm_sem <- bm_sem_path %>%
  image_read()
p_bm_sem_rich <- ggdraw() +
  draw_image(img_bm_sem)
```

@@

```{r}
sem_caption <- paste0(
  "Results of the Structural Equation Models for (A) the determinants of biomass stability and (B) total biomass. Red and green represent respectively negative and positive path coefficients. The path coefficients are standardized coeffients. For convenience, only significant paths (i.e, *P < 0.05*) are shown."

)
```

```{r sem-plot, fig.dim = c(10,10), fig.cap = sem_caption}
#https://tex.stackexchange.com/questions/20496/drawing-different-tikz-shapes-parameterized-by-data-from-a-file?rq=1
#https://tex.stackexchange.com/questions/83888/how-to-plot-data-from-a-csv-file-using-tikz-and-csvsimple

plot_grid(p_stab_sem_rich, p_bm_sem_rich, labels = LETTERS[1:2], rel_widths = c(.50, .5))
```

@@

The average trophic level was the only direct driver of
total biomass ($r_\delta = `r tlvl_on_bm`$) whereas
connectance and species richness had no significant effects (Fig.
\@ref(fig:sem-plot) C). Environmental variables  no direct effects on total
biomass.

Species richness had a indirect positive effect on total biomass through
average trophic level ($`r 0.38*.51`$). The stream size had a total effect on total biomass as strong
as species richness ($r_\delta =`r env_on_bm["log_RC1"]`$), indirectly through average trophic level
and species richness. The avg temperature of the river had an indirect positive
effect on total biomass ($r_\delta = `r env_on_bm["log_RC2"]`$), through
average trophic level and species richness as well. Avg DBO, CV DBO and CV of
river size had no or weak indirect effect on total biomass
(resp. $r_\delta =`r env_on_bm["log_RC3"]`$, $r_\delta =`r env_on_bm["RC5"]`$, $r_\delta =`r env_on_bm["RC4"]`$).

@@

$\overline{CV_{sp}}$ ($r_\delta = `r cv_sp_on_stab`$) decreased biomass stability more than
synchrony ($r_\delta = `r sync_on_stab`$, Figure \@ref(fig:sem-plot)). Species richness had a strong
positive effect on $\overline{CV_{sp}}$ and a strong negative effect on synchrony
whereas average trophic level and connectance had no significant effects on either
$\overline{CV_{sp}}$ nor synchrony. The environmental variables had
no significant direct effects on either $\overline{CV_{sp}}$ nor synchrony.

@@

```{r}
richness_lab <- expression(bold(paste("Total richness (", m^"-2", ") (Log 10)")))
stab_lab <- "Biomass stability (Log 10)"
bm_lab <- expression(bold(paste("Total biomass (", g.m^"-2", ") (Log 10)")))

@@

p_stab_div <- sem_data %>%
  ggplot(aes(x = log_rich_tot_std, y = log_stab_std))+
  geom_point() +
  ylim(lim_log_stab) +
  xlim(lim_log_rich_tot) +
  labs(x = richness_lab, y = stab_lab) +
  geom_line(data = stab_div_pred) +
  geom_ribbon(data = stab_div_pred, aes(ymin = conf.low, ymax = conf.high), fill
  = "grey20", alpha = .2) +
  annotate("text", x = -2.75, y = .75, label = make_label_rsq(rsq = stab_div_sum),
    parse = TRUE)

# Right plot
p_stab_div_troph <- stab_div_troph %>%
  ggplot(aes(x = log_rich_tot_std, y = log_stab_std, color = troph_group)) +
  geom_point() +
  ggplot2::scale_color_manual(
    labels = troph_group_labeller(),
    values = c("blue", "green", "red"),
    name = "Trophic group"
  ) +
  ylim(lim_log_stab) +
  xlim(lim_log_rich_tot) +
  geom_line(data = stab_troph_pred) +
  geom_ribbon(data = stab_troph_pred,
    aes(ymin = conf.low, ymax = conf.high, color = NULL, group = troph_group),
    alpha = .2) +
  annotate("text", x = -2.75, y = .75, label = make_label_rsq(rsq = stab_troph_sum),
    parse = TRUE) +
  #scale_x_log10() + scale_y_log10() +
  labs(x = richness_lab, y = stab_lab)

troph_legend <- get_legend(
  p_stab_div_troph + 
    guides(color = guide_legend(nrow = 1)) +
    theme(legend.position = "bottom")
)

p_stab <- plot_grid(
  p_stab_div,
  p_stab_div_troph + theme(legend.position = "none"),
  ncol = 2, rel_widths = c(.4, .6),
  labels = c("A", "B") 
)
```

@@

p_bm_div <- sem_data %>%
  ggplot(aes(x = log_rich_tot_std, y = log_bm_std))+
  geom_point() +
  ylim(lim_log_bm) +
  xlim(lim_log_rich_tot) +
  labs(x = richness_lab, y = bm_lab) +
  geom_line(data = bm_div_pred) +
  geom_ribbon(data = bm_div_pred, aes(ymin = conf.low, ymax = conf.high), fill
  = "grey20", alpha = .2) +
  annotate("text", x = -2.75, y = -1.0, label = make_label_rsq(rsq = bm_div_sum),
    parse = TRUE)

p_bm_div_troph <- stab_div_troph %>%
  ggplot(aes(x = log_rich_tot_std, y = log_bm_std, color = troph_group)) +
  geom_point() +
  ylim(lim_log_bm) +
  xlim(lim_log_rich_tot) +
  ggplot2::scale_color_manual(
    labels = troph_group_labeller(),
    values = c("blue", "green", "red"),
    name = "Trophic group"
  ) +
  geom_line(data = bm_troph_pred) +
  geom_ribbon(data = bm_troph_pred,
    aes(ymin = conf.low, ymax = conf.high, color = NULL, group = troph_group),
    alpha = .2) +
  annotate("text", x = -2.75, y = -1.0, label = make_label_rsq(rsq = bm_troph_sum),
    parse = TRUE) +
  labs(x = richness_lab, y = bm_lab)

p_bm <- plot_grid(
  p_bm_div,
  p_bm_div_troph + theme(legend.position = "none"),
  ncol = 2,
  rel_widths = c(.4, .6),
  labels = c("C", "D")
)
```
```{r}
troph_caption <- paste0(
  "The effect of species richness on biomass stability
  (upper panel) and total biomass (bottom panel) for the whole community (left
    panel) and the community splited by trophic group (right panel)."
)

```{r psbm, fig.dim = c(10,5), fig.cap = troph_cap}
plot_grid(
  p_stab,
  p_bm,
  troph_legend,
  ncol = 1,
  rel_heights = c(1, 1, .1)
)

@@

**Truc chelou sur l'effet du niveau trophique sur la biomasse et la stabilité,
to check again the computation. Plots look good but the model coefficients does
not go in the same direction**

---
# Total biomass versus species richness
---
Species richness increased total biomass (`r b_co_r`% `r b_ci_r` by 1% species
richness increase). Total biomass was `r b_co_t`% `r b_ci_t` higher in the high
than in the low trophic
group. However, the increase of total biomass with the increase of species
richness was lower in the high than the low trophic group (`r b_co_r_t3`% `r
b_ci_r_t3` and `r b_co_r_t2`% `r b_ci_r_t2` respectively,  by 1% species
richness increase).

@@

We found that species richness increased total biomass as found in salt
marshes, forests [@wu_relationship_2014,@shuaifeng_positive_2018]. But opposite
relation has been found in richness in grasslands [@grace_integrative_2016],
notably because of light competition. Precedent researches in plant communities
showed that the total biomass was primary driven by environmental drivers. In
contrast, we found that the average trophic level was one of the main driver of the
total biomass, together with the stream size and increasing avg temperature
(with decreasing altitude). The effect of species richness is 
goes through its effect on food web assembly and so, avg trophic level.
The indirect positive effect of species richness on total biomass might be due to
resource use complementarity [@cardinale_species_2002].
It was however unexpected that connectance had no effect on total biomass,
because we could imagine that connectance could also improve resource use, but
this might be true only under the assumption that all species have the same
assimilation efficiency.
We found that the effect of stream size, altitude of the river and its avg
temperature on total biomass was only indirect through community structure
(species richness and average trophic level).

@@

The total effect of river size on biomass stability was negative as avg
temperature (and low altitude), indicating that big rivers in plains had a high
negative effect on stability (TO MOVE). We found a mix of both expected and
unexpected results on the effects on environment on community structure. First,
we expected that bigger and low altitude streams had higher species richness
because they might receive more
confluents streams and so more species through immigration.
Stream size increased avg trophic level. An increase in mean trophic position
of piscivorous species in bigger rivers could explain this relation
[@post_ecosystem_2000]. In our network reconstruction, the fish-fish
interactions are more resolved than fish-resource interactions.
@petchey_predicting_2010 predicted that temperature should increase connectance
but we did not find any relationship but temperature (and
low altitude) increased avg trophic level (**Bullshit**). 
CV of enrichment decreased both connectance and avg trophic level, while avg
enrichment (and CV of flow) increased avg trophic level. The latter might by explained by
the fact that enrichment can benefit to top trophic level
[@hulot_functional_2000]. One explanation for the former is that enrichment
variation might destabilize top predators (REF or Bullshit again).

@@

In turn, we found that synchrony and $\overline{CV_{sp}}$ both decreased biomass
stability, as predicted by the formulation of stability based on the CV
[@thibaut_understanding_2013]. However, it was unexpected that the average
population $\overline{CV_{sp}}$ had a stronger negative effect on biomass
stability than synchrony [@bluthgen_land_2016; @olivier_independent_nodate]. We
can advance two hypothesis that can raise this effect. First, the opportunity
for diet differentiation can be low in streams compare to other organisms. Bird
species are for example quite specialized in seed size, flying or soil arthropods or
mammals which gives opportunities for independent dynamics. Plants also can
achieve independent dynamics either spatial segregation or niche
differentiation. Because fish species even piscivorous one are quite
opportunistic in their foraging.
Secondly, in trophic communities, species directly affect each other through
feeding interactions, and so could raise synchrony between species
[@raimondo_interspecific_2004] more than in systems where interactions happen
more in the form of competitive exploitation.
